---
title: "Snps_stats"
author: "Alex Di Genova"
date: "2024-12-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load InStrain data

We use variants from InStrain caller.

```{r loadinstratin}
library(tidyverse)
library(patchwork)
d1=read.table(gzfile("../data/snps/InStrain/is_S15_def_SNVs.tsv.gz"),h=T,sep="\t")

```

We filter for a minimum of read coverage >=5 for each allele

```{r filter1}
# Filter rows where both the reference and variant allele have at least 5 reads
depht=5
filtered_d1 <- d1 %>%
  filter(ref_base %in% c("A", "C", "T", "G") &  # Ensure valid base
           con_base %in% c("A", "C", "T", "G") &
           var_base %in% c("A", "C", "T", "G") &
           (case_when(
              ref_base == "A" ~ A >= depht,
              ref_base == "C" ~ C >= depht,
              ref_base == "T" ~ T >= depht,
              ref_base == "G" ~ G >= depht,
              TRUE ~ FALSE
            )) &
           (case_when(
              var_base == "A" ~ A >= depht,
              var_base == "C" ~ C >= depht,
              var_base == "T" ~ T >= depht,
              var_base == "G" ~ G >= depht,
              TRUE ~ FALSE
            )))

# View the filtered data
head(filtered_d1)

```

We perform some analysis at MAG level, first we add the metadata

```{r maginstrain}

mag=read.table("../data/snps/S15.ctg2bins.txt",h=F)
colnames(mag)=c("MAG","CTG")
filtered_d1=left_join(filtered_d1,mag,by=c("scaffold"="CTG"))
#contig coverage
ccov=read.table("../data/snps/S15_bins.depth.mosdepth.summary.txt",h=T)
ccov=left_join(ccov,mag,by=c("chrom"="CTG"))
#we compute mag coverage
mag_cov=ccov %>% group_by(MAG) %>%summarize(l=sum(length),bases=sum(bases)) %>% mutate(mean_cov=bases/l)

#we load some mag stats from checkM result
mag_qual=read.delim("../data/MAG_SMAG/mags_mining_S15.tsv",header=T)
mag_qual$mag=paste0(mag_qual$mag_name,".",mag_qual$mag_id)
mag_qual=mag_qual %>% select(mag,phylum,class,order,completeness,contamination,strain_heterogeneity,genome_size,gc,N50)

mag_mut_d1=filtered_d1 %>% group_by(MAG) %>% summarise(n=n()) %>% arrange(desc(n))
#we join mag mut and cov
mag_mut_cov_d1=left_join(mag_mut_d1,mag_cov)
mag_mut_cov_d1$mutations_per_megabase <- mag_mut_cov_d1$n / (mag_mut_cov_d1$l / 1e6)
mag_mut_cov_tax_d1=left_join(mag_mut_cov_d1,mag_qual, by=c("MAG"="mag"))
```

We build a linear model to test for association regarding the number of mutations
 
```{r lmodeld1}
a=lm(data=mag_mut_cov_tax_d1,mutations_per_megabase~mean_cov+contamination+gc+strain_heterogeneity+completeness)
summary(a)
```

```{r modelplotd1}
library(RColorBrewer)
# Extract p-values for each predictor
p_values <- summary(a)$coefficients[-1, 4]  # Exclude intercept
predictor_names <- names(p_values)

# Reshape data to long format for faceted plotting
long_data <- mag_mut_cov_tax_d1 %>%
  pivot_longer(
    cols = c(mean_cov, contamination, gc, strain_heterogeneity, completeness),
    names_to = "predictor",
    values_to = "value"
  )
unique_phyla <- unique(long_data$phylum)
phylum_colors <- setNames(scales::hue_pal()(length(unique_phyla)), unique_phyla)


# Define a consistent color palette for phylum
#phylum_colors <- c(
#  "Actinobacteriota" = "#1f77b4", "Proteobacteria" = "#ff7f0e", "Acidobacteriota" = "#2ca02c", 
#  "Firmicutes_E" = "#d62728", "Nitrospirota_A" = "#9467bd", "Dormibacterota" = "#8c564b")


w# Add p-values as a label to the dataset
long_data <- long_data %>%
  mutate(
    p_value_label = case_when(
      predictor == "mean_cov" ~ paste0("p = ", signif(p_values["mean_cov"], 3)),
      predictor == "contamination" ~ paste0("p = ", signif(p_values["contamination"], 3)),
      predictor == "gc" ~ paste0("p = ", signif(p_values["gc"], 3)),
      predictor == "strain_heterogeneity" ~ paste0("p = ", signif(p_values["strain_heterogeneity"], 3)),
      predictor == "completeness" ~ paste0("p = ", signif(p_values["completeness"], 3)),
      TRUE ~ NA_character_
    )
  )

# Plot with facets and p-value annotations
p1=ggplot(long_data, aes(x = value, y = mutations_per_megabase, color=phylum)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  facet_wrap(~ predictor, scales = "free_x") +
  geom_text(data = distinct(long_data, predictor, p_value_label),  # Avoid duplicate labels
            aes(x = Inf, y = Inf, label = p_value_label),
            hjust = 1.1, vjust = 1.5, inherit.aes = FALSE) +
    scale_color_manual(values = phylum_colors) +  # Apply consistent colors
  labs(
    title = "Mutations vs Predictors",
    x = "Predictor Value",
    y = "Mutations per Megabase"
  ) +
  theme_minimal() + theme(legend.position = "bottom")
```


 
The linear model examines how various factors (mean coverage, contamination, GC content, strain heterogeneity, and completeness) influence the number of mutations per megabase.  ~20.4% of the variance in mutations per megabase is explained by the model. The strongest predictor appears to be mean coverage (p = 0.078), suggesting a possible but weak association between sequencing depth and mutation rate. The overall model is not statistically significant, then there is no strong evidence that these predictors collectively explain mutation rates within our dataset.

```{r anovamut}
# Count MAGs per phylum
phylum_counts <- table(mag_mut_cov_tax_d1$phylum)

# Filter phyla with at least 5 MAGs
valid_phyla <- names(phylum_counts[phylum_counts >= 10])
filtered_data <- mag_mut_cov_tax_d1[mag_mut_cov_tax_d1$phylum %in% valid_phyla, ]

anova_model <- aov(mutations_per_megabase ~ phylum, data = filtered_data)
summary(anova_model)

anova_p_value <- summary(anova_model)[[1]]["Pr(>F)"][1, 1]  # Extract p-value
p_value_label <- paste0("ANOVA p = ", signif(anova_p_value, 3))  # Scientific notation

tukey_results_filtered <- TukeyHSD(anova_model)
print(tukey_results_filtered)


p2=ggplot(filtered_data, aes(x = phylum, y = mutations_per_megabase, fill = phylum)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = phylum_colors) +
  geom_text(aes(x = 1.5, y = max(mutations_per_megabase, na.rm = TRUE), 
                label = p_value_label), inherit.aes = FALSE, size = 3) +  # ANOVA p-value annotation
  labs(
    title = "",
    x = "Phylum",
    y = "Mutations per Megabase"
  ) +
  theme_minimal() +
  theme(legend.position = "none")


(p1+p2) + plot_layout(widths = c(2, 1))  + plot_annotation(tag_levels = 'A')
pdf("SNPQC.pdf",width=8)
(p1+p2) + plot_layout(widths = c(2, 1))  + plot_annotation(tag_levels = 'A')
dev.off()

png("SNPQC.png",width = 800, height = 600)
(p1+p2) + plot_layout(widths = c(2, 1))  + plot_annotation(tag_levels = 'A')
dev.off()

```

Figure: SNP quality control and variant calling in MAGs
(A) The linear model assesses the influence of multiple factors—including mean coverage, contamination, GC content, strain heterogeneity, and completeness—on the number of mutations per megabase. The model explains approximately 20.4% of the variance in mutation rates. Among the predictors, mean coverage exhibits the strongest association (p = 0.078), suggesting a possible but weak relationship between sequencing depth and mutation rate. However, the overall model is not statistically significant, indicating that these factors do not collectively provide strong explanatory power for mutation rate variation in our dataset. (B) There is no strong evidence to suggest that the mean number of mutations per megabase differs significantly across the phyla included in the filtered dataset.





## DN/DS analysis

# Global dN/dS per-sample (ANI 95%)

For downstream analysis, we proceeded exploring the results obtained with the caller inStrain (at ANI 95%).

```{r dndsglob}
library(ggbeeswarm)

# Read raw data from dNdScv results
dnds_global <- read.csv("../data/snps/rev02_dnds_global_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
print(dnds_global)

# Plot global dN/dS ratios per sample
snp_plot_global <- dnds_global %>%
  ggplot(aes(x = mle, y = reorder(name, mle), color = name)) +
  geom_pointrange(aes(xmin = cilow, xmax = cihigh), alpha = 0.8) +
  scale_color_manual(values = c("#66c2a5", "#8da0cb", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5"), guide = "none") +
  labs(x = "Global dN/dS ratios", y = "Sample", title = "Global dN/dS ratios (dNdScv)") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5))
snp_plot_global
```

# Global dN/dS per-MAGs (top three phyla)

```{r dndss15}
library(ggpubr)

# Read raw data from dNdScv results
dnds_dndscv <- read.csv("../data/snps/rev02_dnds_sam_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(dnds_dndscv)

# Normalize dN/dS per-gene by contamination
dnds_mags_sam <- dnds_dndscv %>%
  dplyr::mutate(dnds_adj = wmis_cv * (1 - contamination / 100)) %>%
  dplyr::group_by(sample, phylum, bin) %>% #separate by samples
  dplyr::summarise(snps = n(), dnds_sum = sum(dnds_adj), .groups = "drop") %>%
  ungroup() %>%
  dplyr::mutate(dnds_glob = dnds_sum / snps / 8) #divide by total genes/samples
head(dnds_mags_sam)

# Group genes across samples for a merged faceting
dnds_mags_sam_all <- dnds_mags_sam %>%
  dplyr::group_by(phylum, bin) %>%
  dplyr::summarise(dnds = mean(dnds_glob), .groups = "drop") %>%
  dplyr::rename(dnds_glob = dnds) %>%
  dplyr::mutate(sample = "All samples") %>%
  relocate(sample, .before = phylum)
nrow(dnds_mags_sam_all)
dnds_mags_sam_all <- rbind(dnds_mags_sam %>% dplyr::select(sample, phylum, bin, dnds_glob), dnds_mags_sam_all)

# Boxplot MAGs dN/dS ratios
snp_plot_mags <- dnds_mags_sam_all %>%
  dplyr::mutate(sample = factor(sample, levels = c("S1", "S2", "S3", "S5", "S7", "S15", "S17", "S28", "All samples"))) %>% #set order of samples
  dplyr::filter(phylum == "Acidobacteriota" | phylum == "Actinobacteriota" | phylum == "Proteobacteria") %>% #top 3 phyla
  dplyr::mutate(phylum = factor(phylum, levels = c("Proteobacteria", "Acidobacteriota", "Actinobacteriota"))) %>%
  ggplot(aes(x = phylum, y = dnds_glob, color = phylum)) +
  geom_quasirandom(size = 1.5, alpha = 0.85) +
  geom_boxplot(fill = NA, outlier.size = 0, outliers = FALSE) +
  scale_color_manual(values = c("#b15928", "#a6cee3", "#98d277"), guide = "none") +
  #geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  #stat_compare_means(method = "t.test", label = "p.format", vjust = 0.25, size = 2,
  #                   comparisons = list(c("Proteobacteria", "Acidobacteriota"),
  #                                      c("Proteobacteria", "Actinobacteriota"),
  #                                      c("Acidobacteriota", "Actinobacteriota")),
  #                   bracket.size = 0.25, tip.length = 0.01) + #label.y = c(1.05, 1.15, 1.25),
  labs(x = "Phylum", y = "MAG dN/dS ratio", title = "MAGs per-sample") +
  facet_wrap(~ sample) + ylim(c(0,1))+
  theme_minimal() +
  theme(axis.line.x.bottom = element_line(linewidth = 0.5),
        axis.line.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
snp_plot_mags

# Extract bins from S15 according to their equivalence after dRep (dereplication)
dnds_mags_s15 <- dnds_mags_sam %>%
  dplyr::filter(grepl("S15.", bin) |
                bin == "SPAdes-MetaBAT2Refined-S5.34" |
                bin == "SPAdes-MetaBAT2Refined-S3.159_sub" |
                bin == "SPAdes-MetaBAT2Refined-S5.59" |
                bin == "SPAdes-MetaBAT2Refined-S7.123_sub" |
                bin == "SPAdes-MetaBAT2Refined-S17.18" |
                bin == "SPAdes-MetaBAT2Refined-S2.122_sub" |
                bin == "SPAdes-MetaBAT2Refined-S5.72" |
                bin == "SPAdes-MaxBin2Refined-S3.007_sub")
dnds_mags_s15 <- dnds_mags_s15 %>%
  dplyr::mutate(bin = case_when(
    bin == "SPAdes-MetaBAT2Refined-S5.34" ~ "SPAdes-MetaBAT2Refined-S15.45",
    bin == "SPAdes-MetaBAT2Refined-S3.159_sub" ~ "SPAdes-MaxBin2Refined-S15.001_sub",
    bin == "SPAdes-MetaBAT2Refined-S5.59" ~ "SPAdes-MetaBAT2Refined-S15.15_sub",
    bin == "SPAdes-MetaBAT2Refined-S7.123_sub" ~ "SPAdes-MaxBin2Refined-S15.019_sub",
    bin == "SPAdes-MetaBAT2Refined-S17.18" ~ "SPAdes-MetaBAT2Refined-S15.14",
    bin == "SPAdes-MetaBAT2Refined-S2.122_sub" ~ "SPAdes-MetaBAT2Refined-S15.68",
    bin == "SPAdes-MetaBAT2Refined-S5.72" ~ "SPAdes-MetaBAT2Refined-S15.20",
    bin == "SPAdes-MaxBin2Refined-S3.007_sub" ~ "SPAdes-MetaBAT2Refined-S15.18",
    TRUE ~ bin))
head(dnds_mags_s15)

# Load dN/dS for S15 MAGs and add filtered bins
dnds_s15 <- read.csv("../data/snps/dnds_S15_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
dnds_mags_s15 <- dnds_mags_s15 %>% dplyr::filter(sample == "S15")
dim(dnds_mags_s15)
dnds_mags_s15_add <- dnds_s15 %>%
  dplyr::mutate(sample = "S15") %>%
  relocate(sample, .before = phylum) %>%
  dplyr::filter(!bin %in% dnds_mags_s15$bin)
dnds_mags_s15 <- rbind(dnds_mags_s15, dnds_mags_s15_add)
dim(dnds_mags_s15)

# Prepare data with ratios of S15 vs all samples
dnds_mags_s15_corr <- dnds_mags_s15 %>%
  dplyr::select(phylum, bin, dnds_glob) %>%
  dplyr::rename(dnds_all = dnds_glob) %>%
  left_join(dplyr::select(dnds_s15, bin, dnds_glob), by = "bin") %>%
  dplyr::rename(dnds_s15 = dnds_glob)
head(dnds_mags_s15_corr)

# Now, we can correlate the 44 MAGs from S15 across all samples
dnds_s15_corr <- lm(dnds_s15 ~ dnds_all, data = dnds_mags_s15_corr)
#dnds_s15_corr_rs <- summary(dnds_s15_corr)$r.squared
#dnds_s15_corr_pv <- summary(dnds_s15_corr)$coefficients[2, 4]

# Run an additional correlation (p-test)
dnds_s15_cortest <- cor.test(dnds_mags_s15_corr$dnds_s15, dnds_mags_s15_corr$dnds_all,
                             method = "spearman", conf.level = 0.95)
#dnds_s15_cortest

# Plot linear model
snp_plot_mags_lm <- dnds_mags_s15_corr %>%
  ggplot(aes(x = dnds_s15, y = dnds_all, color = phylum)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", linewidth = 0.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  scale_y_continuous(limits = c(0, 0.8)) +
  labs(x = "dN/dS (S15)", y = "dN/dS (S15 across all samples)", title = "dN/dS ratios of S15 MAGs") +
  annotate("text", x = 0.2, y = 0.7, color = "black", size = 5, hjust = 0,
           label = paste0("Spearman's rank correlation = ", round(dnds_s15_cortest$estimate, 3), 
                          "\nP-value = ", signif(dnds_s15_cortest$p.value, 3))) +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5),legend.position = "bottom")
snp_plot_mags_lm


```

Spearman p-test exhibit a correlation coefficient of 0.612 (strong correlation) with a significant p-value of 1.575e-05.


# Intrinsic mining genes under negative selection (NS)

```{r dndsns}
library(ggpubr)

# Read results from dNdScv
dnds_dndscv <- read.csv("../data/snps/rev02_dnds_sam_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Get genes under NS
dnds_mags_nsgenes <- dnds_dndscv %>% dplyr::filter(wmis_cv < 1 & wmis_cv != 0)

# Read list with intrinsic genes extracted from the annotation (GFF file)
snp_cds_genes <- read.csv("../data/snps/selected_cds_metals.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Discard unknown gene names
snp_cds_genes <- snp_cds_genes %>% dplyr::filter(nchar(name) < 14)

# Subset list to those genes under NS (the same `gene_id` could appear in different samples)
dnds_mags_nsgenes_sub <- snp_cds_genes[snp_cds_genes$gene_id %in% dnds_mags_nsgenes$gene_name, ]
dnds_mags_nsgenes_sub <- left_join(dnds_mags_nsgenes_sub, dnds_mags_nsgenes, by = join_by(gene_id == gene_name))

# Separate genes by pathway
fx_snpGene2Path <- function(g) {
  if (grepl("copper", g, ignore.case = TRUE))
    return("Copper")
  if (grepl("iron", g, ignore.case = TRUE))
    return("Iron")
  if (grepl("sulf|soxr", g, ignore.case = TRUE))
    return("Sulfur")
  else
    return("Other")
}
snp_nspath <- vector()
for (i in 1:nrow(dnds_mags_nsgenes_sub)) {
	snp_nspath <- append(snp_nspath, fx_snpGene2Path(dnds_mags_nsgenes_sub$product[i]))
}
dnds_mags_nsgenes_sub$pathway <- snp_nspath
dnds_mags_nsgenes_sub <- dnds_mags_nsgenes_sub %>% dplyr::filter(pathway != "Other")
head(dnds_mags_nsgenes_sub)

# Sort pathways and set colors
dnds_ns_sort <- dnds_mags_nsgenes_sub %>%
  dplyr::group_by(pathway) %>%
  summarise(avg = mean(wmis_cv)) %>%
  arrange(desc(avg))
dnds_ns_sort$color <- c("#ddad3b", "#7f7f7f", "#8c2421")

# Group genes across samples for a merged faceting
dnds_mags_nsgenes_sub_all <- dnds_mags_nsgenes_sub %>% dplyr::mutate(sample = "All samples")
dnds_mags_nsgenes_sub <- rbind(dnds_mags_nsgenes_sub, dnds_mags_nsgenes_sub_all)

# Violin plot of all NS genes by pathway across samples
snp_plot_ns <- dnds_mags_nsgenes_sub %>%
  dplyr::mutate(sample = factor(sample, levels = c("S1", "S2", "S3", "S5", "S7", "S15", "S17", "S28", "All samples"))) %>% #set order of samples
  ggviolin(x = "pathway", y = "wmis_cv", color = "pathway", shape = "pathway", add = "jitter",
           order = dnds_ns_sort$pathway, palette = dnds_ns_sort$color,
           legend = "none", size = 0.5, add.params = list(size = 1, alpha = 0.75)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  #stat_compare_means(method = "t.test", label = "p.signif", vjust = 0.25, size = 2,
  #                  comparisons = list(c("Sulfur", "Iron"), c("Sulfur", "Copper"), c("Iron", "Copper")),
  #                   label.y = c(1.1, 1.2, 1.3), bracket.size = 0.25, tip.length = 0.01) +
  labs(x = "Pathway", y = "dN/dS ratio", title = "Mining genes under NS") +
  facet_wrap(~ sample) +
  theme(strip.background = element_blank()) 
snp_plot_ns
```

The list of intrinsic mining genes (copper, iron, and sulfur proteins) are highly represented with negative selection pressure (dN/dS <1).



```{r mergedndsp}
(snp_plot_global|snp_plot_mags)/(snp_plot_mags_lm|snp_plot_ns) + plot_layout(widths = c(1, 2)) + plot_annotation(tag_levels = 'A')
#library(patchwork)
pdf("Reproducibility_of_seleccion_analysis.pdf",width=12,height = 10)
(snp_plot_global|snp_plot_mags)/(snp_plot_mags_lm|snp_plot_ns) + plot_layout(widths = c(1, 2)) + plot_annotation(tag_levels = 'A')
#(p1+p2) + 
dev.off()

png("Reproducibility_of_seleccion_analysis.png",width=800,height = 600)
(snp_plot_global|snp_plot_mags)/(snp_plot_mags_lm|snp_plot_ns) + plot_layout(widths = c(1, 2)) + plot_annotation(tag_levels = 'A')
#(p1+p2) + 
dev.off()

```
Figure: Evaluation of selection results for S15 across eight mining samples.  
(A) Global dN/dS ratios for S15 and seven additional mining samples confirm strong negative selection, demonstrating consistency across datasets.  
(B) Global dN/dS estimates per MAG and sample further support strong negative selection, indicating that the observed patterns in S15 are also present in other mining samples.  
(C) A Spearman correlation test comparing dN/dS estimations in S15 samples with those obtained from 250 dereplicated MAGs reveals a strong correlation (coefficient = 0.612) with a highly significant p-value of 1.575e-05, reinforcing the robustness of the observed selection patterns.  
(D) Key mining-associated genes involved in copper, iron, and sulfur metabolism are consistently under strong negative selection (dN/dS < 1) in S15 and across other mining samples, highlighting evolutionary constraints in these functional pathways.
