---
title: "Snps_stats"
author: "Alex Di Genova"
date: "2024-12-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load InStrain data

We use variants from InStrain caller.

```{r loadinstratin}
library(tidyverse)
library(patchwork)
d1=read.table(gzfile("../data/snps/InStrain/is_S15_def_SNVs.tsv.gz"),h=T,sep="\t")

```

We filter for a minimum of read coverage >=5 for each allele

```{r filter1}
# Filter rows where both the reference and variant allele have at least 5 reads
depht=5
filtered_d1 <- d1 %>%
  filter(ref_base %in% c("A", "C", "T", "G") &  # Ensure valid base
           con_base %in% c("A", "C", "T", "G") &
           var_base %in% c("A", "C", "T", "G") &
           (case_when(
              ref_base == "A" ~ A >= depht,
              ref_base == "C" ~ C >= depht,
              ref_base == "T" ~ T >= depht,
              ref_base == "G" ~ G >= depht,
              TRUE ~ FALSE
            )) &
           (case_when(
              var_base == "A" ~ A >= depht,
              var_base == "C" ~ C >= depht,
              var_base == "T" ~ T >= depht,
              var_base == "G" ~ G >= depht,
              TRUE ~ FALSE
            )))

# View the filtered data
head(filtered_d1)

```

We perform some analysis at MAG level, first we add the metadata

```{r maginstrain}

mag=read.table("../data/snps/S15.ctg2bins.txt",h=F)
colnames(mag)=c("MAG","CTG")
filtered_d1=left_join(filtered_d1,mag,by=c("scaffold"="CTG"))
#contig coverage
ccov=read.table("../data/snps/S15_bins.depth.mosdepth.summary.txt",h=T)
ccov=left_join(ccov,mag,by=c("chrom"="CTG"))
#we compute mag coverage
mag_cov=ccov %>% group_by(MAG) %>%summarize(l=sum(length),bases=sum(bases)) %>% mutate(mean_cov=bases/l)

#we load some mag stats from checkM result
mag_qual=read.delim("../data/MAG_SMAG/mags_mining_S15.tsv",header=T)
mag_qual$mag=paste0(mag_qual$mag_name,".",mag_qual$mag_id)
mag_qual=mag_qual %>% select(mag,phylum,class,order,completeness,contamination,strain_heterogeneity,genome_size,gc,N50)

mag_mut_d1=filtered_d1 %>% group_by(MAG) %>% summarise(n=n()) %>% arrange(desc(n))
#we join mag mut and cov
mag_mut_cov_d1=left_join(mag_mut_d1,mag_cov)
mag_mut_cov_d1$mutations_per_megabase <- mag_mut_cov_d1$n / (mag_mut_cov_d1$l / 1e6)
mag_mut_cov_tax_d1=left_join(mag_mut_cov_d1,mag_qual, by=c("MAG"="mag"))
```
```{r plotsqc}
#library(patchwork)

data <- data.frame(
  Category = c(
    "Total mapped pairs",
    "Pairs passing max insert size threshold",
    "Pairs passing min insert size threshold",
    "Pairs passing mapQ threshold",
    "Pass min read ANI",
    "Total filtered pairs"
  ),
  Percentage = c(100, 100, 99, 99, 92, 91)
)

# Create the barplot
p3=ggplot(data, aes(x = reorder(Category, -Percentage), y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7, color = "black") +
  coord_flip() +  # Flip the coordinates for a horizontal barplot
  labs(
    title = "Mapping and Filtering Metrics",
    x = "Category",
    y = "Percentage"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Show percentage on the y-axis
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histogram for QUAL
p1=ggplot(filtered_d1, aes(x = var_freq)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Variant Allele Frequency", x = "VAF", y = "Frequency") +
  theme_minimal()

# Histogram for dp_data
p2=ggplot(filtered_d1, aes(x = position_coverage)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of variant read depth (DP)", x = "Depth", y = "Frequency") +
  xlim(c(5,1000))+
  theme_minimal()

p3+p2+p1
```
 
We build a linear model to test for association regarding the number of mutations
 
```{r lmodeld1}
a=lm(data=mag_mut_cov_tax_d1,mutations_per_megabase~mean_cov+contamination+gc+strain_heterogeneity+completeness)
summary(a)
```

```{r modelplotd1}
# Extract p-values for each predictor
p_values <- summary(a)$coefficients[-1, 4]  # Exclude intercept
predictor_names <- names(p_values)

# Reshape data to long format for faceted plotting
long_data <- mag_mut_cov_tax_d1 %>%
  pivot_longer(
    cols = c(mean_cov, contamination, gc, strain_heterogeneity, completeness),
    names_to = "predictor",
    values_to = "value"
  )

# Add p-values as a label to the dataset
long_data <- long_data %>%
  mutate(
    p_value_label = case_when(
      predictor == "mean_cov" ~ paste0("p = ", signif(p_values["mean_cov"], 3)),
      predictor == "contamination" ~ paste0("p = ", signif(p_values["contamination"], 3)),
      predictor == "gc" ~ paste0("p = ", signif(p_values["gc"], 3)),
      predictor == "strain_heterogeneity" ~ paste0("p = ", signif(p_values["strain_heterogeneity"], 3)),
      predictor == "completeness" ~ paste0("p = ", signif(p_values["completeness"], 3)),
      TRUE ~ NA_character_
    )
  )

# Plot with facets and p-value annotations
ggplot(long_data, aes(x = value, y = mutations_per_megabase, color=phylum)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  facet_wrap(~ predictor, scales = "free_x") +
  geom_text(data = distinct(long_data, predictor, p_value_label),  # Avoid duplicate labels
            aes(x = Inf, y = Inf, label = p_value_label),
            hjust = 1.1, vjust = 1.5, inherit.aes = FALSE) +
  labs(
    title = "Response vs Predictors with p-values",
    x = "Predictor Value",
    y = "Mutations per Megabase"
  ) +
  theme_minimal()
```


 
Highly significant (p < 0.001). Each unit increase in mean_cov is associated with an increase of ~4.32 mutations per megabase 

The model explains 31.6% of the variance in mutations_per_megabase.

mean_cov: The only significant predictor (p < 0.001). It has a strong, positive effect on mutation rates, meaning higher sequencing coverage results in more detected mutations. This aligns with the expectation that greater coverage improves variant detection sensitivity.

Other predictors (contamination, gc, strain_heterogeneity, completeness) are not statistically significant (p > 0.05), suggesting they do not have strong or consistent effects on mutation rates in this dataset.

```{r anovamut}
# Count MAGs per phylum
phylum_counts <- table(mag_mut_cov_tax_d1$phylum)

# Filter phyla with at least 5 MAGs
valid_phyla <- names(phylum_counts[phylum_counts >= 10])
filtered_data <- mag_mut_cov_tax_d1[mag_mut_cov_tax_d1$phylum %in% valid_phyla, ]

anova_model <- aov(mutations_per_megabase ~ phylum, data = filtered_data)
summary(anova_model)

tukey_results_filtered <- TukeyHSD(anova_model)
print(tukey_results_filtered)


ggplot(filtered_data, aes(x = phylum, y = mutations_per_megabase, fill = phylum)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Number of Mutations per Megabase by Phylum (Filtered)",
    x = "Phylum",
    y = "Mutations per Megabase"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

There is no strong evidence that the mean number of mutations per megabase differs significantly among the phyla included in the filtered dataset.

# Global dN/dS ratios 

We next compared the global selection pressure ratios of MAGs derived from the custom and custom2 variant calling (inStrain ANI 97% and 98%), calculated with dNdScv according to our methodology.

```{r magsdnds}
# Read dN/dS ratios calculated with calling at ANI of 97% and 98%
dnds_instrain <- read.csv("../data/snps/dnds_inStrain_compare.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Fit a linear model comparing ratios
dnds_instrain_lm <- lm(dnds_is97 ~ dnds_is98, data = dnds_instrain)

# Extract R-squared and p-value
dnds_instrain_lm_rs <- summary(dnds_instrain_lm)$r.squared
dnds_instrain_lm_pv <- summary(dnds_instrain_lm)$coefficients[2, 4]

# Plot model (top three phyla)
dnds_plot_instrain_lm <- dnds_instrain %>%
  ggplot(aes(x = dnds_is98, y = dnds_is97, color = phylum)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 0.5) +
  scale_x_continuous(limits = c(0.2, 1.05)) +
  scale_color_manual(values = c("#98d277", "#a6cee3", "#b15928")) +
  geom_vline(xintercept = 1, linetype = 1, linewidth = 0.25) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.25) +
  labs(x = "dN/dS (ANI 98%)", y = "dN/dS (ANI 97%)", title = "Global dN/dS ratios of MAGs") +
  annotate("text", x = 0.2, y = 0.9, color = "black", size = 5, hjust = 0,
           label = paste0("R² = ", round(dnds_instrain_lm_rs, 3), 
                          "\nP-value = ", signif(dnds_instrain_lm_pv, 3))) +
  theme_minimal()

dnds_plot_instrain_lm
```

The model tends to be more rigorous by using inStrain at ANI of 97% (no MAGs with global positive selection were detected).


# Global dN/dS of S15 MAGs vs all samples

Next, we compared the global ratios of the 44 MAGs from S15 (at default ANI: 95%) with the results of this sample mapped across all samples (eight metagenomes; ANI 97% and 98%).

```{r s15dnds}
#library(ggbeeswarm)
# Read the data from each dNdScv run:
dnds_s15 <- read.csv("../data/snps/dnds_S15_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
dnds_s15_is97 <- read.csv("../data/snps/dnds_S15_merged_is97.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
dnds_s15_is98 <- read.csv("../data/snps/dnds_S15_merged_is98.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Boxplot for each dataset
dnds_plot_1 <- dnds_s15 %>%
  #dplyr::mutate(mag = "mag") %>%
  ggplot(aes(x = phylum, y = dnds_glob, color = phylum)) +
  geom_boxplot(fill = NA, outlier.size = 0) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  #geom_quasirandom(alpha = 0.5, size = 2) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 MAGs (ANI 95%)") +
  guides(color = "none") +
  theme_minimal() +
  theme(axis.line = element_line(size = 0.5),
        axis.ticks.x.bottom = element_line(size = 0.5),
        axis.ticks.y.left = element_line(size = 0.5),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

dnds_plot_2 <- dnds_s15_is97 %>%
  ggplot(aes(x = phylum, y = dnds_glob, color = phylum)) +
  geom_boxplot(fill = NA, outlier.size = 0) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 MAGs (ANI 97%)") +
  guides(color = "none") +
  theme_minimal() +
  theme(axis.line = element_line(size = 0.5),
        axis.ticks.x.bottom = element_line(size = 0.5),
        axis.ticks.y.left = element_line(size = 0.5),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

dnds_plot_3 <- dnds_s15_is98 %>%
  ggplot(aes(x = phylum, y = dnds_glob, color = phylum)) +
  geom_boxplot(fill = NA, outlier.size = 0) +
  geom_jitter(width = 0.1, size = 2, alpha = 0.7) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 MAGs (ANI 98%)") +
  guides(color = "none") +
  theme_minimal() +
  theme(axis.line = element_line(size = 0.5),
        axis.ticks.x.bottom = element_line(size = 0.5),
        axis.ticks.y.left = element_line(size = 0.5),
        axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))

## We merge and show them
dnds_plot_1 + dnds_plot_2 + dnds_plot_3
```

We do observe that the selection pressure tends to increase at higher ANI percentages.


