---
title: "Snps_stats"
author: "Alex Di Genova"
date: "2024-12-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## load data

```{r read_vcf}
library(vcfR)
vcf_file <- "../data/snps/S15_filtered.vcf.gz"  # Replace with your VCF file path
vcf_data <- read.vcfR(vcf_file)

```

## QC mutations

```{r pressure, echo=FALSE}
library(tidyverse)
dp_data <- vcfR::extract.info(vcf_data, element = "DP")  # Replace "DP" with your desired INFO field
dp_data <- as.numeric(dp_data)

fixed_data <- getFIX(vcf_data)
qual <- as.numeric(fixed_data[,"QUAL"])
fixed_data=cbind(fixed_data,dp_data) 
#we create a dataframe and convert some text data to numeric
fixed_data=as.data.frame(fixed_data) %>% mutate(QUAL=as.numeric(QUAL),dp_data=as.numeric(dp_data))

fixed_data=fixed_data %>% select(-FILTER,-ID)
```


Some basic plots for quality control

```{r plots}
library(patchwork)
# Histogram for QUAL
p1=ggplot(fixed_data, aes(x = QUAL)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Distribution of Variant Quality", x = "Quality (phred)", y = "Frequency") +
  theme_minimal()

# Histogram for dp_data
p2=ggplot(fixed_data, aes(x = dp_data)) +
  geom_histogram(binwidth = 5, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of variant read depth (DP)", x = "Depth", y = "Frequency") +
  theme_minimal()

p1+p2
```

# MAGs

We load the MAG information for each contig

```{r rmag}
mag=read.table("../data/snps/S15.ctg2bins.txt",h=F)
colnames(mag)=c("MAG","CTG")
fixed_data=left_join(fixed_data,mag,by=c("CHROM"="CTG"))

#contig coverage
ccov=read.table("../data/snps/S15_bins.depth.mosdepth.summary.txt",h=T)
ccov=left_join(ccov,mag,by=c("chrom"="CTG"))
#we compute mag coverage
mag_cov=ccov %>% group_by(MAG) %>%summarize(l=sum(length),bases=sum(bases)) %>% mutate(mean_cov=bases/l)
#we compute number of mutations per MAG
mag_mut=fixed_data %>% group_by(MAG) %>% summarize(count=n())
#we join mag mut and cov
mag_mut_cov=left_join(mag_mut,mag_cov)
mag_mut_cov$mutations_per_megabase <- mag_mut_cov$count / (mag_mut_cov$l / 1e6)
#we plot mutations per megabase

#we load some mag stats from checkM result
mag_qual=read.delim("../data/MAG_SMAG/mags_mining_S15.tsv",header=T)
mag_qual$mag=paste0(mag_qual$mag_name,".",mag_qual$mag_id)
mag_qual=mag_qual %>% select(mag,phylum,class,order,completeness,contamination,strain_heterogeneity,genome_size,gc,N50)
mag_mut_cov_tax=left_join(mag_mut_cov,mag_qual, by=c("MAG"="mag"))

```

we build a linear model to check if variants can be explained by some variable

```{r lmodel}
a=lm(data=mag_mut_cov_tax,log(mutations_per_megabase)~mean_cov+contamination+gc+strain_heterogeneity+completeness+phylum)
summary(a)

# we plot some interesting associations
# Fit a linear model
model <- lm(log(mutations_per_megabase) ~ strain_heterogeneity, data = mag_mut_cov_tax)

# Extract R-squared and p-value
r_squared <- summary(a)$r.squared
p_value <- summary(a)$coefficients[5, 4]

p1=ggplot(mag_mut_cov_tax, aes(x=strain_heterogeneity,y=log(mutations_per_megabase), color=phylum)) +
  geom_point(size = 3) +  # Add points
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") +  # Add linear model
  labs(
    title = "Mutations per Megabase vs Strain heterogeneity",
    x = "Strain heterogeneity",
    y = "Mutations per Megabase"
  ) +
  annotate("text", x = max(mag_mut_cov_tax$strain_heterogeneity) * 0.7, y = max(log(mag_mut_cov_tax$mutations_per_megabase)), 
           label = paste0("R² = ", round(r_squared, 3), 
                          "\nP-value = ", signif(p_value, 3)), 
           color = "black", size = 5, hjust = 0) +
  theme_minimal()

# Fit a linear model
model <- lm(mutations_per_megabase ~ gc, data = mag_mut_cov_tax)

# Extract R-squared and p-value
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2, 4]

p2=ggplot(mag_mut_cov_tax, aes(x=gc,y=mutations_per_megabase, color=phylum)) +
  geom_point(size = 3) +  # Add points
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") +  # Add linear model
  labs(
    title = "Mutations per Megabase vs GC-content",
    x = "GC-content",
    y = "Mutations per Megabase"
  ) +
  annotate("text", x = max(mag_mut_cov_tax$gc) * 0.7, y = max(mag_mut_cov_tax$mutations_per_megabase), 
           label = paste0("R² = ", round(r_squared, 3), 
                          "\nP-value = ", signif(p_value, 3)), 
           color = "black", size = 5, hjust = 0) +
  theme_minimal()


# Fit a linear model
model <- lm(mutations_per_megabase ~ mean_cov, data = mag_mut_cov_tax)

# Extract R-squared and p-value
r_squared <- summary(model)$r.squared
p_value <- summary(model)$coefficients[2, 4]

p3=ggplot(mag_mut_cov_tax, aes(x=mean_cov,y=mutations_per_megabase, color=phylum)) +
  geom_point(size = 3) +  # Add points
  geom_smooth(method = "lm", se = TRUE, color = "blue", linetype = "dashed") +  # Add linear model
  labs(
    title = "Mutations per Megabase vs Genome Coverage",
    x = "Genome coverage",
    y = "Mutations per Megabase"
  ) +
  annotate("text", x = max(mag_mut_cov_tax$mean_cov) * 0.7, y = max(mag_mut_cov_tax$mutations_per_megabase), 
           label = paste0("R² = ", round(r_squared, 3), 
                          "\nP-value = ", signif(p_value, 3)), 
           color = "black", size = 5, hjust = 0) +
  theme_minimal()


(p1+p2+p3) + plot_layout(guides = "collect") & theme(legend.position = "bottom")

```


Plot by phylum

```{r fplot}
library(ggplot2)

# Plot mutations per megabase by phylum
ggplot(mag_mut_cov_tax, aes(x = phylum, y = log(mutations_per_megabase), fill = phylum)) +
  geom_boxplot() +
  scale_y_log10() +  # Use log10 scale for better visualization if the range is wide
  labs(
    title = "Variants per Megabase by Phylum",
    x = "Phylum",
    y = "Variants per Megabase"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Rotate x-axis labels

```


## InStrain

We use variants from InStrain caller using three ANI filters at 95% (default), 97% (custom), and 98% (custom2). It seem that the custom produce better results.

```{r loadinstratin}
d1=read.table(gzfile("../data/snps/InStrain/is_S15_def_SNVs.tsv.gz"),h=T,sep="\t")
#d1=d1 %>% filter(class == "SNV")
d2=read.table(gzfile("../data/snps/InStrain/is_S15_custom_SNVs.tsv.gz"),h=T,sep="\t")
#d2=d2 %>% filter(class == "SNV")
d3 <- read.table(gzfile("../data/snps/InStrain/is_S15_custom2_SNVs.tsv.gz"),h=T,sep="\t")
```

We filter for a minimum of read coverage >=5 for each allele

```{r filter1}
# Filter rows where both the reference and variant allele have at least 5 reads
filtered_d1 <- d1 %>%
  filter(ref_base %in% c("A", "C", "T", "G") &  # Ensure valid base
           con_base %in% c("A", "C", "T", "G") &
           var_base %in% c("A", "C", "T", "G") &
           (case_when(
              ref_base == "A" ~ A >= 5,
              ref_base == "C" ~ C >= 5,
              ref_base == "T" ~ T >= 5,
              ref_base == "G" ~ G >= 5,
              TRUE ~ FALSE
            )) &
           (case_when(
              var_base == "A" ~ A >= 5,
              var_base == "C" ~ C >= 5,
              var_base == "T" ~ T >= 5,
              var_base == "G" ~ G >= 5,
              TRUE ~ FALSE
            )))

# View the filtered data
head(filtered_d1)

filtered_d1=left_join(filtered_d1,mag,by=c("scaffold"="CTG"))

# ANI of 97%: filter rows where both the reference and variant allele have at least 5 reads
filtered_d2 <- d2 %>%
  filter(ref_base %in% c("A", "C", "T", "G") &  # Ensure valid base
           con_base %in% c("A", "C", "T", "G") &
           var_base %in% c("A", "C", "T", "G") &
           (case_when(
              ref_base == "A" ~ A >= 5,
              ref_base == "C" ~ C >= 5,
              ref_base == "T" ~ T >= 5,
              ref_base == "G" ~ G >= 5,
              TRUE ~ FALSE
            )) &
           (case_when(
              var_base == "A" ~ A >= 5,
              var_base == "C" ~ C >= 5,
              var_base == "T" ~ T >= 5,
              var_base == "G" ~ G >= 5,
              TRUE ~ FALSE
            )))

# View the filtered data
head(filtered_d2)
filtered_d2=left_join(filtered_d2,mag,by=c("scaffold"="CTG"))

# ANI of 98%: filter rows where both the reference and variant allele have at least 5 reads
filtered_d3 <- d3 %>%
  filter(ref_base %in% c("A", "C", "T", "G") &  # Ensure valid base
           con_base %in% c("A", "C", "T", "G") &
           var_base %in% c("A", "C", "T", "G") &
           (case_when(
              ref_base == "A" ~ A >= 5,
              ref_base == "C" ~ C >= 5,
              ref_base == "T" ~ T >= 5,
              ref_base == "G" ~ G >= 5,
              TRUE ~ FALSE)) &
           (case_when(
              var_base == "A" ~ A >= 5,
              var_base == "C" ~ C >= 5,
              var_base == "T" ~ T >= 5,
              var_base == "G" ~ G >= 5,
              TRUE ~ FALSE)))

# View the filtered data
head(filtered_d3)
filtered_d3 <- left_join(filtered_d3, mag, by = c("scaffold" = "CTG"))
```

We perform some analysis at MAG level, first we add the metadata

```{r maginstrain}
mag_mut_d1=filtered_d1 %>% group_by(MAG) %>% summarise(n=n()) %>% arrange(desc(n))
#we join mag mut and cov
mag_mut_cov_d1=left_join(mag_mut_d1,mag_cov)
mag_mut_cov_d1$mutations_per_megabase <- mag_mut_cov_d1$n / (mag_mut_cov_d1$l / 1e6)
mag_mut_cov_tax_d1=left_join(mag_mut_cov_d1,mag_qual, by=c("MAG"="mag"))

#custom 97%
mag_mut_d2=filtered_d2 %>% group_by(MAG) %>% summarise(n=n()) %>% arrange(desc(n))
#we join mag mut and cov
mag_mut_cov_d2=left_join(mag_mut_d2,mag_cov)
mag_mut_cov_d2$mutations_per_megabase <- mag_mut_cov_d2$n / (mag_mut_cov_d2$l / 1e6)
mag_mut_cov_tax_d2=left_join(mag_mut_cov_d2,mag_qual, by=c("MAG"="mag"))

#custom 98%
mag_mut_d3 <- filtered_d3 %>% group_by(MAG) %>% summarise(n=n()) %>% arrange(desc(n))
#we join mag mut and cov
mag_mut_cov_d3 <- left_join(mag_mut_d3, mag_cov)
mag_mut_cov_d3$mutations_per_megabase <- mag_mut_cov_d3$n / (mag_mut_cov_d3$l / 1e6)
mag_mut_cov_tax_d3 <- left_join(mag_mut_cov_d3, mag_qual, by = c("MAG" = "mag"))
```
 
```{r plotsqc}
#library(patchwork)

data <- data.frame(
  Category = c(
    "Total mapped pairs",
    "Pairs passing max insert size threshold",
    "Pairs passing min insert size threshold",
    "Pairs passing mapQ threshold",
    "Pass min read ANI",
    "Total filtered pairs"
  ),
  Percentage = c(100, 100, 99, 99, 92, 91)
)

# Create the barplot
p3=ggplot(data, aes(x = reorder(Category, -Percentage), y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7, color = "black") +
  coord_flip() +  # Flip the coordinates for a horizontal barplot
  labs(
    title = "Mapping and Filtering Metrics",
    x = "Category",
    y = "Percentage"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Show percentage on the y-axis
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histogram for QUAL
p1=ggplot(filtered_d1, aes(x = var_freq)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Variant Allele Frequency", x = "VAF", y = "Frequency") +
  theme_minimal()

# Histogram for dp_data
p2=ggplot(filtered_d1, aes(x = position_coverage)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of variant read depth (DP)", x = "Depth", y = "Frequency") +
  xlim(c(5,1000))+
  theme_minimal()

p3+p2+p1
```
 

```{r plotsqc2}
#library(patchwork)

data <- data.frame(
  Category = c(
    "Total mapped pairs",
    "Pairs passing max insert size threshold",
    "Pairs passing min insert size threshold",
    "Pairs passing mapQ threshold",
    "Pass min read ANI",
    "Total filtered pairs"
  ),
  Percentage = c(100, 100, 99, 99, 88, 87)
)

# Create the barplot
p3=ggplot(data, aes(x = reorder(Category, -Percentage), y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7, color = "black") +
  coord_flip() +  # Flip the coordinates for a horizontal barplot
  labs(
    title = "Mapping and Filtering Metrics",
    x = "Category",
    y = "Percentage"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Show percentage on the y-axis
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histogram for QUAL
p1=ggplot(filtered_d2, aes(x = var_freq)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Variant Allele Frequency", x = "VAF", y = "Frequency") +
  theme_minimal()

# Histogram for dp_data
p2=ggplot(filtered_d2, aes(x = position_coverage)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of variant read depth (DP)", x = "Depth", y = "Frequency") +
  xlim(c(5,1000))+
  theme_minimal()

p3+p2+p1
```


```{r plotsqc3}
data <- data.frame(
  Category = c(
    "Total mapped pairs",
    "Pairs passing max insert size threshold",
    "Pairs passing min insert size threshold",
    "Pairs passing mapQ threshold",
    "Pass min read ANI",
    "Total filtered pairs"
  ),
  Percentage = c(100, 100, 99, 99, 84, 83)
)

# Create the barplot
p3=ggplot(data, aes(x = reorder(Category, -Percentage), y = Percentage, fill = Category)) +
  geom_bar(stat = "identity", width = 0.7, color = "black") +
  coord_flip() +  # Flip the coordinates for a horizontal barplot
  labs(
    title = "Mapping and Filtering Metrics",
    x = "Category",
    y = "Percentage"
  ) +
  scale_y_continuous(labels = scales::percent_format(scale = 1)) +  # Show percentage on the y-axis
  theme_minimal(base_size = 14) +
  theme(
    legend.position = "none",
    axis.text.y = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    plot.title = element_text(size = 16, face = "bold")
  )

# Histogram for QUAL
p1=ggplot(filtered_d3, aes(x = var_freq)) +
  geom_histogram(binwidth = 0.05, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Variant Allele Frequency", x = "VAF", y = "Frequency") +
  theme_minimal()

# Histogram for dp_data
p2=ggplot(filtered_d3, aes(x = position_coverage)) +
  geom_histogram(binwidth = 10, fill = "orange", color = "black", alpha = 0.7) +
  labs(title = "Distribution of variant read depth (DP)", x = "Depth", y = "Frequency") +
  xlim(c(5,1000))+
  theme_minimal()

p3+p2+p1
```

We build a linear model to test for association regarding the number of mutations
 
```{r lmodeld1}
a=lm(data=mag_mut_cov_tax_d1,mutations_per_megabase~mean_cov+contamination+gc+strain_heterogeneity+completeness)
summary(a)
```

```{r modelplotd1}
# Extract p-values for each predictor
p_values <- summary(a)$coefficients[-1, 4]  # Exclude intercept
predictor_names <- names(p_values)

# Reshape data to long format for faceted plotting
long_data <- mag_mut_cov_tax_d1 %>%
  pivot_longer(
    cols = c(mean_cov, contamination, gc, strain_heterogeneity, completeness),
    names_to = "predictor",
    values_to = "value"
  )

# Add p-values as a label to the dataset
long_data <- long_data %>%
  mutate(
    p_value_label = case_when(
      predictor == "mean_cov" ~ paste0("p = ", signif(p_values["mean_cov"], 3)),
      predictor == "contamination" ~ paste0("p = ", signif(p_values["contamination"], 3)),
      predictor == "gc" ~ paste0("p = ", signif(p_values["gc"], 3)),
      predictor == "strain_heterogeneity" ~ paste0("p = ", signif(p_values["strain_heterogeneity"], 3)),
      predictor == "completeness" ~ paste0("p = ", signif(p_values["completeness"], 3)),
      TRUE ~ NA_character_
    )
  )

# Plot with facets and p-value annotations
ggplot(long_data, aes(x = value, y = mutations_per_megabase, color=phylum)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  facet_wrap(~ predictor, scales = "free_x") +
  geom_text(data = distinct(long_data, predictor, p_value_label),  # Avoid duplicate labels
            aes(x = Inf, y = Inf, label = p_value_label),
            hjust = 1.1, vjust = 1.5, inherit.aes = FALSE) +
  labs(
    title = "Response vs Predictors with p-values",
    x = "Predictor Value",
    y = "Mutations per Megabase"
  ) +
  theme_minimal()
```


 
```{r lmodeld2}
a=lm(data=mag_mut_cov_tax_d2,mutations_per_megabase~mean_cov+contamination+gc+strain_heterogeneity+completeness)
summary(a)
```


```{r modelplotd2}
# Extract p-values for each predictor
p_values <- summary(a)$coefficients[-1, 4]  # Exclude intercept
predictor_names <- names(p_values)

# Reshape data to long format for faceted plotting
long_data <- mag_mut_cov_tax_d2 %>%
  pivot_longer(
    cols = c(mean_cov, contamination, gc, strain_heterogeneity, completeness),
    names_to = "predictor",
    values_to = "value"
  )

# Add p-values as a label to the dataset
long_data <- long_data %>%
  mutate(
    p_value_label = case_when(
      predictor == "mean_cov" ~ paste0("p = ", signif(p_values["mean_cov"], 3)),
      predictor == "contamination" ~ paste0("p = ", signif(p_values["contamination"], 3)),
      predictor == "gc" ~ paste0("p = ", signif(p_values["gc"], 3)),
      predictor == "strain_heterogeneity" ~ paste0("p = ", signif(p_values["strain_heterogeneity"], 3)),
      predictor == "completeness" ~ paste0("p = ", signif(p_values["completeness"], 3)),
      TRUE ~ NA_character_
    )
  )

# Plot with facets and p-value annotations
ggplot(long_data, aes(x = value, y = mutations_per_megabase, color=phylum)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  facet_wrap(~ predictor, scales = "free_x") +
  geom_text(data = distinct(long_data, predictor, p_value_label),  # Avoid duplicate labels
            aes(x = Inf, y = Inf, label = p_value_label),
            hjust = 1.1, vjust = 1.5, inherit.aes = FALSE) +
  labs(
    title = "Response vs Predictors with p-values",
    x = "Predictor Value",
    y = "Mutations per Megabase"
  ) +
  theme_minimal()
```

```{r lmodeld3}
a <- lm(data=mag_mut_cov_tax_d3,mutations_per_megabase~mean_cov+contamination+gc+strain_heterogeneity+completeness)
summary(a)
```

```{r modelplotd3}
# Extract p-values for each predictor
p_values <- summary(a)$coefficients[-1, 4]  # Exclude intercept
predictor_names <- names(p_values)

# Reshape data to long format for faceted plotting
long_data <- mag_mut_cov_tax_d3 %>%
  pivot_longer(
    cols = c(mean_cov, contamination, gc, strain_heterogeneity, completeness),
    names_to = "predictor",
    values_to = "value"
  )

# Add p-values as a label to the dataset
long_data <- long_data %>%
  mutate(
    p_value_label = case_when(
      predictor == "mean_cov" ~ paste0("p = ", signif(p_values["mean_cov"], 3)),
      predictor == "contamination" ~ paste0("p = ", signif(p_values["contamination"], 3)),
      predictor == "gc" ~ paste0("p = ", signif(p_values["gc"], 3)),
      predictor == "strain_heterogeneity" ~ paste0("p = ", signif(p_values["strain_heterogeneity"], 3)),
      predictor == "completeness" ~ paste0("p = ", signif(p_values["completeness"], 3)),
      TRUE ~ NA_character_
    )
  )

# Plot with facets and p-value annotations
ggplot(long_data, aes(x = value, y = mutations_per_megabase, color=phylum)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = TRUE, color = "blue") +
  facet_wrap(~ predictor, scales = "free_x") +
  geom_text(data = distinct(long_data, predictor, p_value_label),  # Avoid duplicate labels
            aes(x = Inf, y = Inf, label = p_value_label),
            hjust = 1.1, vjust = 1.5, inherit.aes = FALSE) +
  labs(
    title = "Response vs Predictors with p-values",
    x = "Predictor Value",
    y = "Mutations per Megabase"
  ) +
  theme_minimal()
```

Highly significant (p < 0.001). Each unit increase in mean_cov is associated with an increase of ~4.32 mutations per megabase 

The model explains 31.6% of the variance in mutations_per_megabase.

mean_cov: The only significant predictor (p < 0.001). It has a strong, positive effect on mutation rates, meaning higher sequencing coverage results in more detected mutations. This aligns with the expectation that greater coverage improves variant detection sensitivity.

Other predictors (contamination, gc, strain_heterogeneity, completeness) are not statistically significant (p > 0.05), suggesting they do not have strong or consistent effects on mutation rates in this dataset.

```{r anovamut}
# Count MAGs per phylum
phylum_counts <- table(mag_mut_cov_tax_d2$phylum)

# Filter phyla with at least 5 MAGs
valid_phyla <- names(phylum_counts[phylum_counts >= 5])
filtered_data <- mag_mut_cov_tax_d2[mag_mut_cov_tax_d2$phylum %in% valid_phyla, ]

anova_model <- aov(mutations_per_megabase ~ phylum, data = filtered_data)
summary(anova_model)

tukey_results_filtered <- TukeyHSD(anova_model)
print(tukey_results_filtered)


ggplot(filtered_data, aes(x = phylum, y = mutations_per_megabase, fill = phylum)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.2, alpha = 0.5) +
  labs(
    title = "Number of Mutations per Megabase by Phylum (Filtered)",
    x = "Phylum",
    y = "Mutations per Megabase"
  ) +
  theme_minimal() +
  theme(legend.position = "none")

```

There is no strong evidence that the mean number of mutations per megabase differs significantly among the phyla included in the filtered dataset.

# Correlation between callers (bcftools and inStrain)

```{r s15callers}
# Read data
dnds_comp <- read.csv("../data/snps/dnds_S15_bcf_vs_is.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(dnds_comp)

# Fit a linear model
dnds_comp_lm <- lm(dnds_bcf ~ dnds_is95, data = dnds_comp)

# Extract R-squared and p-value
dnds_comp_lm_rs <- summary(dnds_comp_lm)$r.squared
dnds_comp_lm_pv <- summary(dnds_comp_lm)$coefficients[2, 4]

# Plot model
dnds_plot_comp <- dnds_comp %>%
  ggplot(aes(x = dnds_bcf, y = dnds_is95, color = phylum)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 0.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  geom_vline(xintercept = 1, linetype = 1, linewidth = 0.25) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.25) +
  labs(x = "dN/dS (bcftools)", y = "dN/dS (inStrain ANI 95%)", title = "S15 comparison of dN/dS") +
  annotate("text", x = 0.2, y = 0.9, color = "black", size = 5, hjust = 0,
           label = paste0("R² = ", round(dnds_comp_lm_rs, 3), 
                          "\nP-value = ", signif(dnds_comp_lm_pv, 3))) +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5))
dnds_plot_comp
```

We observe a lack of correlation between both callers (R² = 0.049, p-value = 0.15).

# S15 MAGs dN/dS ratios (ANI 97-98%)

We next compared the global selection pressure ratios of MAGs derived from the custom and custom2 variant calling (inStrain ANI 97% and 98%), calculated with dNdScv according to our methodology.

```{r magsdnds}
# Read dN/dS ratios calculated with calling at ANI of 97% and 98%
dnds_instrain <- read.csv("../data/snps/dnds_inStrain_compare.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Fit a linear model comparing ratios
dnds_instrain_lm <- lm(dnds_is97 ~ dnds_is98, data = dnds_instrain)

# Extract R-squared and p-value
dnds_instrain_lm_rs <- summary(dnds_instrain_lm)$r.squared
dnds_instrain_lm_pv <- summary(dnds_instrain_lm)$coefficients[2, 4]

# Plot model (top three phyla)
dnds_plot_instrain_lm <- dnds_instrain %>%
  ggplot(aes(x = dnds_is98, y = dnds_is97, color = phylum)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = TRUE, color = "blue", linewidth = 0.5) +
  scale_x_continuous(limits = c(0.2, 1.05)) +
  scale_color_manual(values = c("#a6cee3", "#98d277", "#b15928")) +
  geom_vline(xintercept = 1, linetype = 1, linewidth = 0.25) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.25) +
  labs(x = "dN/dS (ANI 98%)", y = "dN/dS (ANI 97%)", title = "S15 MAGs dN/dS") +
  annotate("text", x = 0.2, y = 0.9, color = "black", size = 5, hjust = 0,
           label = paste0("R² = ", round(dnds_instrain_lm_rs, 3), 
                          "\nP-value = ", signif(dnds_instrain_lm_pv, 3))) +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5))
dnds_plot_instrain_lm
```

The model tends to be more rigorous by using inStrain at ANI of 97% (no MAGs with global positive selection were detected).

# S15 MAGs dN/dS across all samples (ANI 95-98%)

Next, we compared the global ratios of the 44 MAGs from S15 (at default ANI: 95%) with the results of this sample mapped across all samples (eight metagenomes; at selected ANI of 95%, 97% and 98%).

```{r s15dnds}
library(ggbeeswarm)

# Read global ratios of the 44 MAGs from S15
dnds_s15 <- read.csv("../data/snps/dnds_S15_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(dnds_s15)

# Boxplot for S15
dnds_plot_1 <- dnds_s15 %>%
  dplyr::mutate(mag = "") %>%
  ggplot(aes(x = mag, y = dnds_glob, color = phylum)) +
  geom_boxplot(fill = NA, color = "black", outlier.size = 0) +
  geom_quasirandom(alpha = 0.75, size = 2.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 MAGs (ANI 95%)") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5))
dnds_plot_1

# Read the selection results of S15 mapped across all samples:
dnds_s15_is95 <- read.csv("../data/snps/dnds_S15_merged_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
dnds_s15_is97 <- read.csv("../data/snps/dnds_S15_merged_is97.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
dnds_s15_is98 <- read.csv("../data/snps/dnds_S15_merged_is98.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)

# Boxplot across samples/ANI:
dnds_plot_2 <- dnds_s15_is95 %>%
  dplyr::mutate(mag = "") %>%
  ggplot(aes(x = mag, y = dnds, color = phylum)) +
  geom_boxplot(fill = NA, color = "black", outlier.size = 0) +
  geom_quasirandom(alpha = 0.75, size = 2.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 at ANI 95%") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5))

dnds_plot_3 <- dnds_s15_is97 %>%
  dplyr::mutate(mag = "") %>%
  ggplot(aes(x = mag, y = dnds, color = phylum)) +
  geom_boxplot(fill = NA, color = "black", outlier.size = 0) +
  geom_quasirandom(alpha = 0.75, size = 2.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 at ANI 97%") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5))

dnds_plot_4 <- dnds_s15_is98 %>%
  dplyr::mutate(mag = "") %>%
  ggplot(aes(x = mag, y = dnds, color = phylum)) +
  geom_boxplot(fill = NA, color = "black", outlier.size = 0) +
  geom_quasirandom(alpha = 0.75, size = 2.5) +
  scale_color_manual(values = c("#A6CEE3", "#98D277", "#F16667", "#FE982C", "#7D54A5", "#B15928")) +
  labs(x = "", y = "Global dN/dS ratios", title = "S15 at ANI 98%") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5))

# We merge and show them
(dnds_plot_2 + dnds_plot_3 + dnds_plot_4) + plot_layout(guides = "collect")
```

First, we do observe that mapping the S15 sample alone gives global dN/dS ratios below 1 (i.e., MAGs evolve under purifying selection). This selection pressure leads to an increase when mapping the S15 MAGs across all samples (after dereplication), and at higher ANI percentages (97-98%).

# Global dN/dS per-sample (ANI 95%)

For downstream analysis, we proceeded exploring the results obtained with the caller inStrain (at ANI 95%).

```{r dndsglob}
# Read data
dnds_global <- read.csv("../data/snps/rev02_dnds_global_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
print(dnds_global)

# Plot global dN/dS ratios per sample
snp_plot_global <- dnds_global %>%
  ggplot(aes(x = mle, y = reorder(name, mle), color = name)) +
  geom_pointrange(aes(xmin = cilow, xmax = cihigh), alpha = 0.8) +
  scale_color_manual(values = c("#66c2a5", "#8da0cb", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5", "#66c2a5"), guide = "none") +
  labs(x = "Global dN/dS ratios", y = "Sample", title = "S15 per-sample") +
  theme_minimal() +
  theme(axis.line = element_line(linewidth = 0.5),
        axis.ticks.y.left = element_line(linewidth = 0.5),
        axis.ticks.x.bottom = element_line(linewidth = 0.5))
snp_plot_global
```

# Global dN/dS per-MAGs (top three phyla)

```{r dndss15}
library(ggpubr)

# Read data
dnds_mags <- read.csv("../data/snps/rev02_dnds_S15_samples_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(dnds_mags)

# Boxplot MAGs dN/dS ratios
snp_plot_mags <- dnds_mags %>%
  dplyr::mutate(sample = factor(sample, levels = c("S1", "S2", "S3", "S5", "S7", "S15", "S17", "S28"))) %>% #set order
  dplyr::filter(phylum == "Acidobacteriota" | phylum == "Actinobacteriota" | phylum == "Proteobacteria") %>% #top 3 phyla
  dplyr::mutate(phylum = factor(phylum, levels = c("Proteobacteria", "Acidobacteriota", "Actinobacteriota"))) %>%
  ggplot(aes(x = phylum, y = dnds_glob, color = phylum)) +
  geom_quasirandom(size = 1.5, alpha = 0.85) +
  geom_boxplot(fill = NA, outlier.size = 0, outliers = FALSE) +
  scale_color_manual(values = c("#b15928", "#a6cee3", "#98d277"), guide = "none") +
  stat_compare_means(method = "t.test", label = "p.format", vjust = 0.25, size = 2,
                     comparisons = list(c("Proteobacteria", "Acidobacteriota"),
                                        c("Proteobacteria", "Actinobacteriota"),
                                        c("Acidobacteriota", "Actinobacteriota")), #c(2.4, 2.5, 2.6)
                     label.y = c(1.1, 1.2, 1.3), bracket.size = 0.25, tip.length = 0.01) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  labs(x = "Phylum", y = "dN/dS ratio", title = "S15 MAGs per-sample") +
  facet_wrap(~ sample, scales = "free_y") +
  theme_minimal() +
  theme(axis.line.x.bottom = element_line(size = 0.5),
        axis.line.y.left = element_line(size = 0.5),
        axis.ticks.x.bottom = element_line(size = 0.5),
        axis.ticks.y.left = element_line(size = 0.5),
        axis.text.x = element_text(angle = 45, hjust = 1))
snp_plot_mags
```

# Intrinsic mining genes under negative selection (NS)

```{r dndsns}
library(ggpubr)

# Read data
dnds_ns <- read.csv("../data/snps/rev02_genes_ns_is95.txt", sep = "\t", header = TRUE, stringsAsFactors = FALSE)
head(dnds_ns)

# Replicate violin across samples
# Sort pathways and set colors to plot
dnds_ns_sort <- dnds_ns %>%
  dplyr::group_by(pathway) %>%
  summarise(avg = mean(wmis_cv)) %>%
  arrange(desc(avg))
dnds_ns_sort$color <- c("#ddad3b", "#7f7f7f", "#8c2421")

# Violin plot of all NS genes by pathway
snp_plot_ns <- dnds_ns %>%
  dplyr::mutate(sample = factor(sample, levels = c("S1", "S2", "S3", "S5", "S7", "S15", "S17", "S28"))) %>% #set order
  ggviolin(x = "pathway", y = "wmis_cv", color = "pathway", shape = "pathway", add = "jitter",
           order = dnds_ns_sort$pathway, palette = dnds_ns_sort$color,
           legend = "none", size = 0.5, add.params = list(size = 1, alpha = 0.75)) +
  geom_hline(yintercept = 1, linetype = 1, linewidth = 0.2) +
  stat_compare_means(method = "t.test", label = "p.signif", vjust = 0.25, size = 2,
                     comparisons = list(c("Sulfur", "Iron"), c("Sulfur", "Copper"), c("Iron", "Copper")),
                     label.y = c(1.1, 1.2, 1.3), bracket.size = 0.25, tip.length = 0.01) +
  labs(x = "Pathway", y = "dN/dS ratio", title = "Mining genes under NS") +
  facet_wrap(~ sample) +
  theme(strip.background = element_blank())
snp_plot_ns
```
